### 2010-10-07 Thesis/Chapter11.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\subsubsection*{Hollow structure modelling.}
We proposed a framework for real-time modelling of thin anatomical structures. We first designed a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist. We improved and extended a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation to allow large displacements (but small deformations). Once combined with an in-plane membrane formulation we obtained an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies. To the best of our knowledge, this co-rotational shell FEM is the first description of a shell finite element method in the field of medical simulation. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. Indeed, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved. Therefore, we suggested a simple method to mesh a surface with curved shell elements. The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells. We also showed how contacts and interactions with the curved surfaces of shells can be handled using the same function.

@@

The TLED algorithm offers an accurate modelling for soft tissues and its GPU implementation using CUDA is very fast. Hence, real-time computations can be achieved without any trade-off on the accuracy of the deformation. One of the features allowing such an efficiency is the use of an explicit time integration scheme. If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step. This critical time step depends on two kind of criterion: the size of the elements and the material?s properties (Young's modulus and Poisson's ratio). Therefore, meshing of anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements. Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time. Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an intensification of the simulation's frequency to enforce real-time. In some applications, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance. However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.

@@

We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy. The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects. Nevertheless, the use of a linear strain measure limits the range of application to small deformations only. Moreover, the constitutive law was chosen to be the simple (linear) Hookean law, which it is known not to describe accurately the relationship between strain and stress in soft tissues. However, those potential changes are substantial and their implementations are not straightforward. In addition, it is not clear from which method the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated). Our shell FEM could also be improved by implementing the formulation on GPU. Although an optimal parallel implementation is always challenging, our method is highly parallelisable and there is no major obstacle to an efficient GPU implementation.

@@

One of the strongest constraint in medical simulation is the demand for real-time computations. This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photorealistic rendering etc.). In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures. The lack of accuracy induced by these simplifications is of great concern to us. In this PhD, a part ot the solution was to accelerate the computations by using GPU. While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the panacea. Firstly, a standard is yet to be imposed in GPGPU. While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless. In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance. Secondly, GPU programming is not simple to master. Although we have seen many improvements in flexibility of programmation over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code. As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least. But most importantly, GPU acceleration is only interesting if the code is highly parallelisable. The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties. Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps. Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance).

@@

Earlier this year, NVIDIA released a new GPU architecture called \emph{Fermi}. Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, 3 times the amount of shared memory, etc. In theory, this series of improvement should boost the performance of the second and third TLED implementations (see section \ref{chap6:secondTLED}) and we should investigate and compare the efficiency of the three designs on this new hardware.

@@

Although the meshing of complex surfaces with curved shell elements was not the heart of this PhD, our naïve approach could be enhanced by focusing on element quality. One simple solution would be to add a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could account for another PhD.

>>>>

\subsubsection*{Hollow structure modelling.}
We proposed a framework for real-time modelling of thin anatomical structures. We first designed a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist. We improved and extended a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation to allow large displacements (but small deformations). Once combined with an in-plane membrane formulation we obtained an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies. To the best of our knowledge, this co-rotational shell FEM is the first description of a shell finite element method in the field of medical simulation. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. Indeed, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved. Therefore, we suggested a simple method to mesh a surface with curved shell elements. The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells. We also showed how contacts and interactions with the curved surfaces of shells can be detected and handled using the same function.

@@

The TLED algorithm offers an accurate modelling for soft tissues and its GPU implementation using CUDA is very fast. Hence, real-time computations can be achieved without any trade-off on the accuracy of the deformation. One of the features allowing such an efficiency is the use of an explicit time integration scheme. If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step. This critical time step depends on two kinds of criteria: the size of the elements and the material's properties (Young's modulus and Poisson's ratio). Therefore, the meshing of the anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements. Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time. Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an intensification of the simulation's frequency to enforce real-time. In some applications, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance. However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.

@@

We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy. The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects. Nevertheless, the use of a linear strain measure limits the range of application to small deformations only. Moreover, the constitutive law was chosen to be the simple (linear) Hookean law, which it is known not to describe accurately the relationship between strain and stress in soft tissues. However, those potential changes are substantial and their implementations are not straightforward. In addition, it is not clear from which method the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true mathematical shell theory (where membrane and bending energies are not separated). Our shell FEM could also be improved by implementing the formulation on GPU. Although an optimal parallel implementation is always challenging, our method is highly parallelisable and there is no major obstacle to an efficient GPU implementation.

@@

One of the strongest constraint in medical simulation is the demand for real-time computations. This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photorealistic rendering etc.). In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures. The lack of accuracy induced by these simplifications is of great concern to us. In this PhD, a part ot the solution was to accelerate the computations by using the GPU. While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the panacea. Firstly, a standard is yet to be imposed in GPGPU. While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless. In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance (up to twice as slow). Secondly, GPU programming is not simple to master. Although we have seen many improvements in flexibility of programmation over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code. As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least. But most importantly, GPU acceleration is only interesting if the code is highly parallelisable. The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties. Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps. Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance).

@@

Earlier this year, NVIDIA released a new GPU architecture called \emph{Fermi}. Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, three times the amount of shared memory, etc. In theory, this series of improvement should boost the performance of the second and third TLED implementations (see section \ref{chap6:secondTLED}) and we should investigate and compare the efficiency of the three designs on this new hardware.

@@

Although the meshing of complex surfaces with curved shell elements was not the heart of this PhD, our naïve approach could be enhanced by focusing on element quality. One simple solution would be to add a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could certainly account for another PhD.

### 2010-10-07 Thesis/Chapter3.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

It is crucial to keep in mind that approximations occur at different stages during finite element analysis. The division of the whole domain into finite elements may not be exact (see \fig{chap3:fig-errorWithSubdivision}), introducing error in the domain being modelled. The second stage is when element equations are derived. As mentionned earlier, the unknowns of the problem are approximated using the idea that any continuous function can be representated by a linear combination of known functions and undetermined coefficients. Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element. There are a few types of approach for establishing these equations but, without going into details, the mathematical foundations of all these approaches are energy principles or the \emph{weighted residual methods} (see Appendix~\ref{appendix2}), which both lead to integrals during the process. Therefore, the second stage creates two sources of error: the representation of the solution by a linear combination of functions and the evaluation of the integrals. Finally, errors are introduced when solving the assembled system of algebraic equations. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision1.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision2.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision3.pdf}}
\caption[Error introduced by the division into elements]{The division of the domain (a) into elements is not always perfect (b). The discretisation error may be reduced by using smaller elements (c). However, this improvement is obtained to the detriment of computational efficiency as the number of algebraic equations to solve will increase.}
\label{chap3:fig-errorWithSubdivision}
\end{figure}

>>>>

It is crucial to keep in mind that approximations occur at different stages during finite element analysis. The division of the whole domain into finite elements may not be exact (see \fig{chap3:fig-errorWithSubdivision}), introducing error in the domain being modelled. The second stage is when element equations are derived. As mentionned earlier, the unknowns of the problem are approximated using the idea that any continuous function can be representated by a linear combination of known functions and undetermined coefficients. Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element. There are a few types of approach for establishing these equations but, without going into details, the mathematical foundations of all these approaches are energy principles or the \emph{weighted residual methods} (see Appendix~\ref{appendix1}), which both lead to integrals during the process. Therefore, the second stage creates two sources of error: the representation of the solution by a linear combination of functions and the evaluation of the integrals. Finally, errors are introduced when solving the assembled system of algebraic equations. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision1.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision2.pdf}}
\hfill 
\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision3.pdf}}
\caption[Error introduced by the division into elements]{The division of the domain (a) into elements is not always perfect (b). The discretisation error may be reduced by using smaller elements (c). However, this improvement is obtained to the detriment of computational efficiency as the number of algebraic equations to solve will increase.}
\label{chap3:fig-errorWithSubdivision}
\end{figure}

### 2010-10-07 Thesis/Chapter4.tex
- Update 6 October -
Besides acknowledgements, the draft is finished!



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@182 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\ON
The challenges faced by the field of medical simulation were reviewed in details in the introductory chapter (see chapter \ref{chap1}). Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators. In addition, this objective has to be achieved within heavy time constraints. Indeed, although each step of a simulation alone (deformation of organs, collision detection, contacts processing or visualisation) is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Of particular interest to us is the accurate modelling of the deformation of anatomical structures. The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation. Very often simplifications have to be made. Since medical simulation may find applications in various areas (such as medical training, patient-specific planning or per-operative guidance), the requirements in terms of accuracy as well as time constraints may be different. Consequently, a range of models has been developed over time to suit all usages.

@@

Although modal analysis significantly accelerates the simulation, it generates noticeable artifacts when applied to large deformations due to the step of linearisation. It can produce unnatural results with deformations of large amplitude. \cite{Choi05} proposed to overcome this limitation by taking the rotational component of the deformation into account at each node. \cite{Barbic05} make use of modal analysis to build a deformation basis from modal derivatives. It allows them to simulate large deformations of solid objects with non-linear material.

@@

\cite{Mosegaard05a,Mosegaard05b} were probably the first to implement a mass-spring model on GPU in medical simulation. For the first time, the whole simulation of was performed on the GPU (physics, interaction and visualisation) and they used their framework in a congenital cardiac surgery simulation. They compared two implementations: the first one explicitly defined the connectivity between springs using a connectivity texture while the second one determines implicitly determines at runtime that springs are connected when they are neighbours. They were able to compute 188 iterations per second on a $42\,745$ particule model of a pig heart. Depending on the size of the model and the type of implementation, the authors achieved speed-ups between 10 and 30 $ \times $ over their CPU implementation.

Taking advantage of the GPU to reach a substantial speed-up is usually not straightforward. One must understand the limitations inherent in its design and devise algorithms accordingly. With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model.

\TODO{Other GPU-based mass-spring implementations?}

>>>>

\ON
The challenges faced by the field of medical simulation were reviewed in the introductory chapter (see chapter \ref{chap1}). Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators. In addition, this objective has to be achieved within heavy time constraints. Indeed, although each step of a simulation alone (deformation of organs, collision detection, contacts processing or visualisation) is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Of particular interest to us is the accurate modelling of the deformation of anatomical structures. The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation. Very often simplifications have to be made. Since medical simulation may find applications in various areas (such as medical training, patient-specific planning or per-operative guidance), the requirements in terms of accuracy as well as time constraints may be different. Consequently, a range of models has been developed over time to suit all usages.

@@

Although modal analysis significantly accelerates the simulation, it generates noticeable artifacts when applied to large deformations due to the step of linearisation. It can produce unnatural results with deformations of large amplitude. \cite{Choi05a} proposed to overcome this limitation by taking the rotational component of the deformation into account at each node. \cite{Barbic05} make use of modal analysis to build a deformation basis from modal derivatives. It allows them to simulate large deformations of solid objects with non-linear material.

@@

\cite{Mosegaard05a,Mosegaard05b} were probably the first to implement a mass-spring model on GPU in medical simulation. For the first time, the whole simulation of was performed on the GPU (physics, interaction and visualisation) and they used their framework in a congenital cardiac surgery simulation. They compared two implementations: the first one explicitly defines the connectivity between springs using a connectivity texture while the second one implicitly determines at runtime that springs are connected when they are neighbours. They were able to compute 188 iterations per second on a $42\,745$ particule model of a pig heart. Depending on the size of the model and the type of implementation, the authors achieved speed-ups between 10 and 30 $ \times $ over their CPU implementation.

Taking advantage of the GPU to reach a substantial speed-up is usually not straightforward. One must understand the limitations inherent in its design and devise algorithms accordingly. With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model. \ON In 2008, \citeauthor{Rasmusson08} investigated performance and optimisation strategies between the two GPU languages OpenGL and CUDA for their mass-spring-damper model. More recently, \cite{Leon10} presented a GPU implementation of a mass-spring model with a nearly $80\times$ speedup over serial execution and $20\times$ speedup over CPU-based parallel execution. \OFF

### 2010-10-07 Thesis/Chapter6.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
The previous chapter introduced the TLED algorithm in details as well as enhancements with a viscoelastic and anisotropic formulation. If the TLED algorithm had already been implemented on GPU by \cite{Taylor07b}, restrictions due to using a graphics-based API and the limited hardware possibilities of GPUs necessitated reformulation of the force summation in their implementation. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general computation on GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the aim to achieve a more straightforward and efficient implementation. The aim was also to enhance the GPU-based implementation of the TLED with a viscoelastic and anisotropic formulation. This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA. By providing an efficient and accurate non-linear FEM for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators. This work has led to three publications \citep{Comas2008,Taylor2008,Taylor2009}.
\end{shortAbstract}

@@

Interface with a GPU has traditionally been via a graphic application programming interface (API) such as \citetalias{Board05} or \citetalias{DirectX10}. OpenGL is an open standard API that provides a number of functions for the rendering of 2D and 3D graphics and is available on most modern operating systems including but not limited to Windows, Mac OS X and Linux. It was initially designed by Silicon Graphics Inc. and now managed by the technology consortium Khronos group. DirectX is a proprietary API that provides functions to render three dimensional graphics, and uses hardware acceleration if it is available on the graphics card. It was designed by Microsoft for use on the Windows platform.

@@

	\subsection{Second implementation: a better use of shared memory}

@@

\TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}

@@

\TODO{If time, try simple implementation using atomic writes with floats on Geforce GTX 480}

>>>>

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
The previous chapter introduced the TLED algorithm in details as well as enhancements with a viscoelastic and anisotropic formulation. If the TLED algorithm had already been implemented on GPU by \cite{Taylor07b}, restrictions due to using a graphics-based API and the limited hardware possibilities of GPUs necessitated reformulation of the force summation in their implementation. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general computation on GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the aim to achieve a more straightforward and efficient implementation. The aim was also to enhance the GPU-based implementation of the TLED with a viscoelastic and anisotropic formulation. This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA. This work has led to three publications \citep{Comas2008,Taylor2008,Taylor2009}.
\end{shortAbstract}

@@

Interface with a GPU has traditionally been via a graphic application programming interface (API) such as \citetalias{OpenGL} or \citeauthor{DirectX}. OpenGL is an open standard API that provides a number of functions for the rendering of 2D and 3D graphics and is available on most modern operating systems including but not limited to Windows, Mac OS X and Linux. It was initially designed by Silicon Graphics Inc. and now managed by the technology consortium Khronos group. DirectX is a proprietary API that provides functions to render three dimensional graphics, and uses hardware acceleration if it is available on the graphics card. It was designed by Microsoft for use on the Windows platform.

@@

	\subsection{Second implementation: a better use of shared memory}	\label{chap6:secondTLED}

@@

% \TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}

@@

% \TODO{If time, try simple implementation using atomic writes with floats on Geforce GTX 480}

### 2010-10-07 Thesis/Chapter7.tex
- Update 2 October -
A few changes in shell part + part of conclusion




git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@180 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Modelling the deformation of hollow objects in real-time}
\label{chap7}
\begin{shortAbstract}
The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging. In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs (like brain, liver, prostate etc.). But there is also a need to simulate the deformation of hollow organs, whose volume is negligible compared to their surface area, such as colon, blood vessels, stomach etc. The solid models previously mentioned are not appropriate and specific approaches are required for these thin anatomical structures. In this chapter, we will review diverse techniques applied for the simulation of thin objects. These different approaches may be grouped into three main categories: (1) mass-spring models, (2) approaches which derive a bending energy from geometrical considerations and (3) methods based on the equations of continuum mechanics. Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible. 
\end{shortAbstract}

@@

To improve the computational efficiency, \cite{Hutchinson96} presented a mechanism for adaptively refining the network of mass-springs to concentrate efforts only where it is needed. When potential inaccuracies are detected, the object is locally refined in the affected region, and the simulation may carry on. \cite{Oshita01} used a sparse triangular mesh and an interpolation to generate a dense mesh.

@@

Among the geometry-based models, there is the noticeable work of \cite{Grinspun03} who introduced a discrete shell model. They designed a simple shell model for triangle meshes by geometrically deriving membrane and bending energies. They applied geometric operators over piecewise-linear surfaces to define a discrete constitutive model, which results in simpler expressions than tensorial derivations from shell theory. They applied their method to simulate the deformation of common objects like beams, sheets of paper or hats (see \fig{chap7:fig-hat}). 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=8cm]{chapter7/hat.pdf}
\caption[Comparison between real footage and simulation of a hat dropped on a table]{Comparison between real footage and simulation of a hat dropped on a table. Image courtesy of \cite{Grinspun03}.}
\label{chap7:fig-hat}
\end{center}
\end{figure}

@@

We seek to propose a solution for simulating, in real-time, the deformation of thin anatomical structures. Since \cite{Black91} showed that formulations with only membrane energy were likely to produce significant errors in the stress distribution, this model is required to feature bending. Ideally, this solution would be versatile and allow us to simulate thin objects with various shapes and material properties with good accuracy. In this regard, methods derived from continuum mechanics are the most appropriate. In some situations, a thin plate model would not correctly capture the deformation of some objects and therefore a shell formulation was retained for more flexibility. To the best of our knowledge, no real-time shell finite element method was presented in the field of medical simulation. \TODO{need to triple-check this...}

>>>>

\chapter{Modelling the deformation of hollow objects in real-time}
\label{chap7}
\begin{shortAbstract}
The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging. In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs (like brain, liver, prostate etc.). But there is also a need to simulate the deformation of hollow organs, whose volume is negligible compared to their surface area, such as colon, blood vessels, stomach etc. The solid models previously mentioned are not appropriate and specific approaches are required for these thin anatomical structures. In this chapter, we will review diverse techniques applied for the simulation of thin objects. These different approaches may be grouped into three main categories: (1) models, (2) approaches which derive a bending energy from geometrical considerations and (3) methods based on the equations of continuum mechanics. Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible. 
\end{shortAbstract}

@@

To improve the computational efficiency, \cite{Hutchinson96} presented a mechanism for adaptively refining the network of mass-springs to concentrate efforts only where it is needed. When potential inaccuracies are detected, the object is locally refined in the affected region, and the simulation may carry on. \cite{Oshita01} used a sparse triangular mesh and an interpolation to generate a dense mesh. And \cite{Georgii05} made use of GPU acceleration and compared two different implementations for modelling cloth patches.

@@

Among the geometry-based models, there is the noticeable work of \cite{Grinspun03} who introduced a discrete shell model. They designed a simple shell model for triangle meshes by geometrically deriving membrane and bending energies. They applied geometric operators over piecewise-linear surfaces to define a discrete constitutive model, which results in simpler expressions than tensorial derivations from shell theory. They applied their method to simulate the deformation of common objects like beams, sheets of paper or hats (see \fig{chap7:fig-hat}). 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=9cm]{chapter7/hat.pdf}
\caption[Comparison between real footage and simulation of a hat dropped on a table]{Comparison between real footage and simulation of a hat dropped on a table. Image courtesy of \cite{Grinspun03}.}
\label{chap7:fig-hat}
\end{center}
\end{figure}

@@

We seek to propose a solution for simulating, in real-time, the deformation of thin anatomical structures. Since \cite{Black91} showed that formulations with only membrane energy were likely to produce significant errors in the stress distribution, this model is required to feature bending. Ideally, this solution would be versatile and allow us to simulate thin objects with various shapes and material properties with good accuracy. In this regard, methods derived from continuum mechanics are the most appropriate. In some situations, a thin plate model would not correctly capture the deformation of some objects and therefore a shell formulation was retained for more flexibility. To the best of our knowledge, no real-time shell finite element method was presented in the field of medical simulation.

### 2010-10-07 Thesis/Chapter9.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

	\subsection{Inversion of matrix $ \mathbf{C} $}
\cite{Przemieniecki85} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
\TODO{add condition of inversion... but need the book :)}

Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles). We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki85}.

\TODO{add figure with the two different coordinate system}

>>>>

%	\subsection{Inversion of matrix $ \mathbf{C} $}
%\cite{Przemieniecki85} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
%\TODO{add condition of inversion... but need the book :)}
%
%Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles). We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki85}. 
%
%\TODO{add figure with the two different coordinate system}

### 2010-10-07 Thesis/Resume.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Bien que les deux modèles de tissus sont basés sur la mécanique continue pour une meilleure précision, ils sont tous les deux capables de simulation la déformation d'organes en temps réel. Enfin, leur implémentation dans l'environnement open source SOFA rendra publique ces deux modèles afin de participer à l'amélioration du réalisme des simulateurs médicaux.

>>>>

Bien que les deux modèles de tissus soient basés sur la mécanique continue pour une meilleure précision, ils sont tous les deux capables de simuler la déformation d'organes en temps réel. Enfin, leur implémentation dans l'environnement open source SOFA permettra la diffusion de ces deux modèles afin de participer à l'amélioration du réalisme des simulateurs médicaux.
